---
title: "Data Visualization Lab"
subtitle: |
  
  Introduction to Data Journalism (JOUR 377)
author: "Mariam Cosmos"
pagetitle: "Mariam Cosmos"
date: today

format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
    link-external-newwindow: true

execute:
  warning: false
---

::: {.callout-tip icon="false"}
## GitHub Repo Link

[Mariam Cosmos's GitHub Repo](https://github.com/mejcosmos/Findings_FactChecking_Lab.git)
:::

:::

## Part 1: Fact-check a news story

This dataset was sourced frm the Chicago Finance Department. The data was used in a Chicago Tribune story, "Downtown bike- and bus-lane ticketing program generates more than 3,500 warnings and violations in its first weeks" by Sarah Freishtat, on the Smart Streets pilot program.

To reproduce the analysis, I imported the Chicago Finance Department Smart Streets dataset into R and standardized the column names using janitor::clean_names(). I converted the date column to a Date format so I could reliably filter by time period. I then filtered the dataset to the window referenced in the story (Nov. 6 through Dec. 13, 2024) and counted the number of rows to measure total enforcement actions. To distinguish warnings from tickets, I created two logical variables based on whether the violation_description field contained the word “WARNING.” I repeated this filtering process for the first weeks of December (Dec. 1–Dec. 13) to isolate ticket counts and then used keyword matching within the violation description to categorize ticket types. All totals were produced by counting rows after applying these filters.

```{r}
#| label: load-packages-data
#| echo: false 

# Packages
library(readxl)
library(dplyr)
library(lubridate)
library(janitor)
library(stringr)

# load data 
library(readxl)
factcheck_dataset <- FOIA_KIEFER_WBEZ_Smart_Streets_data_122024 <- read_excel("DATA/FOIA_KIEFER_WBEZ_Smart_Streets data_122024.xlsx")

# clean names and parse date/time
df <- factcheck_dataset |>
  clean_names() |>
  mutate(
    date = as.Date(date)
    #time = suppressWarnings(hms::as_hms(time))
  )

# check of columns present
# names(df)

df_window <- df |>
  filter(date >= as.Date("2024-11-06") &
         date <= as.Date("2024-12-13"))

# Classify WARNING vs TICKET
df_window <- df_window |>
  mutate(
    is_warning = str_detect(str_to_upper(violation_description), "WARNING"),
    is_ticket  = !is_warning
  )


```

#### Fact-check 1:

-   "Between Nov. 6 and Dec. 13, the city issued some 3,577 warnings and tickets to parking scofflaws under the new Smart Streets pilot program."

```{r}
#| label: total-actions
#| echo: true 


# fACT 1
fact1_total_actions <- nrow(df_window)
fact1_total_actions

```

While the article states that the city issued some 3,577 warning and tickets, my analysis showed that 3,584 warnings and tickets were issues. However, in class, I remember you mentioning that the actual total was 3,585. I wonder why in my analysis I got the number 3,584 instead?

#### Fact-check 2:

-   "But the city did issue 13 tickets in the first weeks of December, the data show, which would have been given for repeat violations."

```{r}
#| label: dec-tickets-first-weeks
#| echo: true 

dec_start <- as.Date("2024-12-01")

df_dec_first_weeks <- df_window |>
  filter(date >= dec_start & date <= as.Date("2024-12-13"))

fact2_tickets_dec_first_weeks <- df_dec_first_weeks |>
  filter(is_ticket) |>
  nrow()

fact2_tickets_dec_first_weeks


```

I chose December 1 - 13 as the first weeks of Deceomber. As such, my analysis confirmed that 13 tickets were issued.

#### Fact-check 3:

-   "Even so, the city issued at least 3,564 warnings in the first weeks of the program. Of the tickets issued, eight were for bus lane infractions and five were for bike lane violations, the data show."

```{r}
#| label: warnings-in-window
#| echo: true 

fact3_warnings_in_window <- df_window |>
  filter(is_warning) |>
  nrow()

fact3_warnings_in_window

```

```{r}
#| label: tickets-dec 
#| echo: true 


tickets_dec <- df_dec_first_weeks |>
  filter(is_ticket) |>
  mutate(
    ticket_type = case_when(
      str_detect(str_to_upper(violation_description), "BUS\\s*LANE") ~ "bus_lane",
      str_detect(str_to_upper(violation_description), "BIKE|BICYCLE") ~ "bike_lane",
      TRUE ~ "other"
    )
  )

ticket_type_counts <- tickets_dec |>
  count(ticket_type, name = "n") |>
  arrange(desc(n))

fact3_bus_lane_tickets  <- tickets_dec %>% filter(ticket_type == "bus_lane")  %>% nrow()
fact3_bike_lane_tickets <- tickets_dec %>% filter(ticket_type == "bike_lane") %>% nrow()

fact3_bus_lane_tickets

fact3_bike_lane_tickets

```

Based on my analysis, the main discrepancies appeared in the first and second fact-checks. The first being, the article stating that 3,577 warnings and tickets were issued, whereas I got 3,584. Furthermore, the second fact check, where my dataset produced 3,571 warnings compared with the article’s “at least 3,564.” These differences are small. It could stem from slight date-range interpretation differences, late data updates, or record corrections after publication rather than an analytical error. All other ticket subtype counts, including eight bus-lane and five bike-lane violations, aligned with the story, and I was able to verify most of the claims made in the article using the available dataset.

:::

:::

## Part 2: Extend your fact-check to another data set

This dataset was sourced frm the Chicago Finance Department, and the data was used in a Streetsblog Chicago article, "Stay in your lane or pay the fine" by John Greenfield, on the Smart Streets pilot program.

#### Fact-check 1:

-   "The CTA said that from from Nov 14 to Dec 31, 2025, there were over 2,200 warnings and violations from bus-mounted enforcement alone. Here is the data from both CDOT/DOF vehicle-mounted cameras and CTA bus-mounted cameras for that period."

```{r}
#| label: street-data-cta-bus
#| echo: true 

library(readxl)
library(dplyr)
library(lubridate)
library(janitor)
library(stringr)

smart_streets_data <- read_excel(
  "DATA_PART2/FOIA_Keifer_Smart_Streets_Data_01052026.xlsx"
) |> 
  clean_names()

```

```{r}
#| label: street-data-cta-bus-1
#| echo: true 

smart_streets_data <- smart_streets_data |>
  mutate(issue_day = as.Date(issue_date))

date_window <- smart_streets_data |>
  filter(issue_day >= as.Date("2025-11-14"),
         issue_day <= as.Date("2025-12-31"))

cta_bus <- date_window |>
  filter(!str_detect(speed_camera_id, "^FI"),
         !str_detect(speed_camera_id, "^DT"))

nrow(cta_bus)


```

#### Fact-check 2:

-   "Although most initial enforcement were no-cost warnings, the bus cameras captured 2,204 incidents, 617 of which came with a fine, dictated by violation type."

```{r}
#| label: fine-incidents
#| echo: true 

# set the date window
date_window <- smart_streets_data |>
  mutate(issue_day = as.Date(issue_date)) |>
  filter(issue_day >= as.Date("2025-11-14"),
         issue_day <= as.Date("2025-12-31"))

# CTA bus cameras 
cta_bus <- date_window |>
  filter(!str_detect(speed_camera_id, "^FI"),
         !str_detect(speed_camera_id, "^DT"))

# fined count check
sum(cta_bus$fine_level_1 > 0, na.rm = TRUE)


```

To isolate CTA bus-mounted enforcement, I grouped records by camera ID prefix and found that numeric-only IDs accounted for 2,203 incidents during the period, as opposed to the reported 2,204 incidents in the article. Similarly, my analysis showed that 615 of the incidents came with a fine, dictated by violation type.

#### Fact-check 3:

-   "Overall, the Smart Streets program total has issued 44,390 warnings and violations. To reiterate, this is only with eight DOF vehicles and 6 buses equipped with the technology, indicating how common this really is. The intent is to dissuade this behavior for the benefit of all users, primarily the most vulnerable like pedestrians and cyclists."

```{r}
#| label: total-warnings-violations
#| echo: true 

nrow(smart_streets_data)

```

Furthermore, the article states that the Smart Streets program total has issued 44,390 warnings and violations, my analysis showed that there were 51,165 warnings and violations.

When I examined how many of those incidents resulted in a monetary penalty by summing rows where fine_level_1 was greater than zero, my analysis produced 615 fined incidents compared with the reported 617.

The only substantial discrepancy appeared when verifying the article’s claim that the Smart Streets program had issued 44,390 total warnings and violations overall. Counting all rows in the dataset produced 51,165 records. Unlike the one- or two-record gaps above, this suggests the article may have relied on an earlier dataset version.

While I do not have access to the crash data nor the figures that include cost of buses and other operational cost, if I did have access to these data sets, it would be interesting to fact check the following claims made in the article:

-   "Reducing conflicts between people on foot, bikes, and buses, and in delivery and passenger vehicles, is not easy in tight spaces. The pilot area is home to many businesses, homes, and amenities. With so much activity, it's not surprising the area sees over 12,000 crashes a year. A renewed focus on safety can’t bring back the four vulnerable road users who were fatally struck in the CBD in 2025. But it can pave the way for a safer future."

-   "The cost to deploy the system on the first six buses is \$315,852 including equipment and ongoing services; the contract includes the option to purchase up to 94 additional cameras," said a March 2025 CTA news release, previewing the launch later that year. Currently, the transit agency is not retaining the revenue. Other cities with the program at larger scale have revenue sharing. this would allow the CTA to use ticket fees to implement more dedicated bus lanes, and make other transit improvements.

-   "This data is well visualized with the Chicago Crash Dashboard, created by Michael McLean from Strong Towns Chicago. He estimated that downtown collisions resulted in roughly \$240 million in economic costs."

:::

:::

# Part 3: Exchange lab work

I shared my lab work with Dina Drogin.

Across both parts, my fact-checking process focused on reproducing each article’s quantitative claims directly from the underlying Smart Streets datasets using R. I began by importing the spreadsheets from the Chicago Finance Department, cleaning column names, and converting date fields so I could reliably filter to the exact time windows referenced in each story. From there, I verified totals by counting rows after applying simple filters, and I distinguished warnings from tickets by searching for the word “WARNING” in the violation descriptions. For more detailed claims, such as bus-lane versus bike-lane tickets or bus-mounted versus vehicle-mounted cameras, I used keyword matching and camera ID prefixes to categorize records before producing counts.

Most of the figures in both articles were reproducible or very close to the published numbers, with only minor differences of one to several records in some early totals. These small gaps are not especially concerning. The only substantial discrepancy appeared in the second article’s overall program total, where the dataset contained noticeably more records than reported, suggesting the journalist may have relied on an earlier version of the dataset.

:::

:::

# Part 4: Best practices

Based on the labs and readings, the following are three best practices I'll keep in mind when fact-checking data-driven work for publication:

1.  I should verify the time frame and data version, since small differences in dates or updated datasets can change totals.
2.  Do not filter and do the analysis on the raw dataset, instead make a copy.
3.  While I am most familiar with R, I should work on my data analysis and fact-checking on spreadsheets or Excel. I do not think R was the best platform to do this kind of fact-checking because I think using R made it overly complicated.

:::
